---
title: "Getting Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{css, echo=FALSE}
body .main-container {
max-width: 1280px !important;
width: 1280px !important;
}
body {
max-width: 1280px !important;
}
```

```{r setup, include=FALSE}
library(oncoRegimens)

knitr::opts_chunk$set(echo = TRUE)

```

ARTEMIS provides the user with a tool that can generate sets of alignment scores, using a temporally modified local Needleman-Wunsch / Smith-Waterman algorithm, for comparing regimens and patient drug records derived from electronic health records.

## Connecting ARTEMIS to the OMOP CDM

ARTEMIS interacts directly with the Observational Medical Outcomes Partnership (OMOP) Common Data Model (CDM), a convenient and widely used format for storing observational health data derived from electronic health records. For users of ARTEMIS with access to data already stored in an OMOP CDM compliant datasource, drug records can be easily and automatically mapped to the drug record strings required by ARTEMIS.

A toy dataset is included in the ARTEMIS package, and will be used throughout this short vignette.

We'll use a [duckdb](https://duckdb.org/) static file database for this purpose.


```{r Setup Con}
con <- DBI::dbConnect(duckdb::duckdb(), here::here("data/vignette/Artemis_Vignette_5.3_0.8.duckdb"))

cdm <- CDMConnector::cdm_from_con(con = con,
                                  cdm_schema = "main",
                                  write_schema = "main")

```

## Selecting patients (A cohort)

Before running ARTEMIS, a cohort json object is required. Cohorts are a fundamental building block for observational data analysis and each cohort is simply constructed by selecting a set of persons or patients who satisfy a set of inclusion criteria, such as a specific condition or a specific drug exposure. Users of ARTERMIS are expected to utilise a cohort of cancer patients, but any cohort format is accepted. More information on cohorts can be found within [The Book of OHDSI](https://ohdsi.github.io/TheBookOfOhdsi/Cohorts.html).

Again, a toy cohort example is included for the purposes of this vignette, selecting all patients within our database. We use the package CDMConnector to both read our cohort description file (In json format) into R and to construct a cdm object armed with this cohort set.

```{r Setup Cohorts}
validPatients <- CDMConnector::readCohortSet(path = here::here("exampleCohort/vignette/"))
cohortName <- "vignette"

cdm <- CDMConnector::generate_cohort_set(cdm, 
                                         cohort_set = validPatients, 
                                         name = cohortName,
                                         overwrite = TRUE)

```

## Loading Regimen and Drug data

To successfully run ARTEMIS, a user must provide both a list of valid drugs and a set of regimens which they would like to search for among patient records. The valid drug list must include all drugs found within any regimens within the set, and should also include any relevant drugs that are not found in regimens but are still clinically relevant. 

In this vignette example, we will explore three pain suppression regimens - two paracetamol and ibuprofen monotherapies and a combination therapy of both drugs. As such, we include paracetamol and ibuprofen within our valid drugs. We also include aspirin, as patients who take aspirin in combination with either of these two drugs, or alone, are likely to be on a different pain suppression regimen which we must account for. We choose not to include unrelated drugs, such as warfarin, as these are likely only to interfere with regimen discovery and are not included in a typical pain suppression regimen.

The choice of which drugs are considered valid can have large effects on the final score of all regimens.

In the case of cancer patients, a large set of valid drugs has been collected from the HemOnc reference resource, which can be loaded with the loadDrugs() function and a set of regimens derived from the same source may be loaded via loadRegimens() In this vignette, we will use 2 custom .csv files to load our data.



The minimum valid format for a valid drugs dataframe:

| name      | valid_concept_id |
| ----------- | ----------- |
| acetaminophen      | 1125315       |
| ibuprofen   | 1177480        |
| aspirin   | 1112807        |



The minimum valid format for a regimen dataframe:

| regName      | shortString |
| ----------- | ----------- |
| Acetaminophen Monotherapy | 1.acetaminophen;7.acetaminophen;                         |
| Ibuprofen Monotherapy     | 1.ibuprofen;7.ibuprofen;                                 |
| Acetaminophen & Ibuprofen | 1.acetaminophen;0.ibuprofen;7.acetaminophen;0.ibuprofen; |



By creating such files and importing them as .csv's, custom regimens and sets of drugs may be easily supplied by the user with which to utilise the ARTEMIS implementation of the TSW algorithm.

```{r Load Data}
validDrugs <- read.csv(here::here("data/vignette/example_drugMap_Valid.csv"))
regimens <- read.csv(here::here("data/vignette/example_regimens.csv"))
```

## Creating a string dataframe

The temporal smith-waterman (TSW) algorithm requires that input drug records take the form of individual strings, with both drug and temporal information stored within the string. The function stringDF_from_cdm() allows us to generate such strings from a given EHR. The format of each patient's total drug record thus takes the form:

[Days since last exposure].[Drug exposure];[Days since last exposure][Drug exposure];...;

Drug exposures which occur on the same day are listed as having 0 days since last exposure. The specific order of exposures that occur on the same day will not effect the TSW score, as these will be dynamically reordered during alignment.

This dataframe may then be filtered to include only patients with a number of drug exposures above a certain cut-off value. First, we must pull our CohortSet into R via CDMConnector.

```{r Create and Filter StringDF}
con_df <- getCohortSet(cdm = cdm, json = validPatients, name = cohortName)

stringDF <- stringDF_from_cdm(con_df = con_df, writeOut = F, validDrugs = validDrugs)

stringDF <- stringDF %>% filter_stringDF(min = 20)

stringDF
```

## Running the algorithm

Once a set of patients have been summarized into a strinDF data frame, and all relevant regimen data are loaded, alignment via the TSW algorithm can commence.

```{r Running the alignment}
options(width=800)

output_all <- stringDF %>% generateRawAlignments(regimens = regimens,
                                                 g = 0.4,
                                                 Tfac = 0.5,
                                                 verbose = 0,
                                                 mem = -1,
                                                 removeOverlap = -1,
                                                 method = "PropDiff")

output_all[c(4,5,6,107,108,109,269,268,267),c(1,3,4,7,8,11,12)]

```

Our output file contains the local score of any passing regimens at any point throughout each patient's drug record. All results are calculated exhaustively, but only those with a total score (Raw score / Number of aligned pairs) greater than 0.6 are returned.

An output file may contain a large number of low-scoring and overlapping regimens, especially when many similar sets of regimens are used as input. A post-processing step may be performed to reduce this unnecessary complexity by condensing and combining regimens into discrete, non-overlapping regimen eras.

These processed alignments may then be visualized and compared to their source drug record in order to validate the results of an alignment visually.

```{r Post-processing/Plotting, fig.width=10, fig.align="center", warning=FALSE}

output_processed <- processAlignments(rawOutput = output_all, regimenCombine = 28, returnDrugs = TRUE)

plotProcesssed(output_processed[output_processed$personID=="3",])

plotProcesssed(output_processed[output_processed$personID=="135",])

plotProcesssed(output_processed[output_processed$personID=="115",])

```

